---
title: "An Extended Family Approach to Income and Wealth Inequality in the U.S."
author: "Names"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
keywords: ["put keywords here"]
bibliography: bibliography_clan.bib
csl: council-of-science-editors-brackets.csl
always_allow_html: true
output:
  pdf_document:
    number_sections: false
    extra_dependencies:
      - float
      - graphicx
      - codehigh
      - ulem
      - siunitx
  word_document:
    number_sections: false
header-includes:
  - \usepackage{tabularray}
  - \UseTblrLibrary{booktabs}
  - \UseTblrLibrary{siunitx}
  - \newcommand{\tinytableTabularrayUnderline}[1]{\underline{#1}}
  - \newcommand{\tinytableTabularrayStrikeout}[1]{\sout{#1}}
  - \NewTableCommand{\tinytableDefineColor}[3]{\definecolor{#1}{#2}{#3}}
  - \usepackage{float}
  - \floatplacement{figure}{H}
  - \usepackage{caption}
  - \captionsetup[table]{labelformat=empty}
knit: (function(input, ...) {
  rmarkdown::render(input, ..., output_format = "pdf_document",
    output_file = here::here("10_draft", "output", "write.pdf"));
  rmarkdown::render(input, ..., output_format = "word_document",
    output_file = here::here("10_draft", "output", "write.docx"))
  })
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = FALSE,
  warning = FALSE,
  message = FALSE,
  tidy    = FALSE
)
options(width = 10000)  # prevent R from wrapping long strings in output
library(here)
library(dplyr)
library(flextable)
library(grid)
library(tidyr)
library(purrr)
library(readr)
library(ggplot2)
library(cowplot)
library(officer)
library(grid)
library(gridExtra)
library(tidyverse)
library(ineq)

# Formatting helpers for inline numbers
fmt2 <- function(x) format(round(x, 2), nsmall = 2)
fmt1 <- function(x) format(round(x, 1), nsmall = 1)
fmt0 <- function(x) format(round(x, 0), big.mark = ",")
fmtp <- function(x) paste0(format(round(x, 2), nsmall = 2), "%")

# Helper for adding notes to figures in .docx and .pdf
print_note <- function(txt) {
  if (knitr::is_latex_output()) {
    cat("\n\n\\begingroup\\itshape\\small\\par\n", txt, "\n\\endgroup\n\n")
  } else {
    # Word: blank lines ensure a new paragraph; txt wrapped in markdown italic
    cat(paste0("\n\n*", txt, "*\n\n"))
  }
}
```

```{r load}
SAVE_FILES <- TRUE
source(here::here("9_figures", "src", "all_figures.R"))
```

```{r scalars}
# All scalars are computed in figures.R and are available in the environment.
```

# Abstract

Most macro-level studies of inequality treat households as the primary unit of analysis.
As a result, we know far less about inequality across extended families. Building on
research highlighting the importance of intergenerational family ties, we examine income
and wealth inequality in the United States using both households and intergenerational
extended families, which we term clans, as units of analysis. Using data from the Panel
Study of Income Dynamics (PSID), we calculate Gini coefficients for income and wealth at
the household and clan levels from 1969 to 2023. We find that inequality is substantially
lower when measured at the clan rather than the household level. The income Gini declines
by `r fmt2(inc_gini_diff)` points on average (from `r fmt2(avg_inc_hh)` to `r fmt2(avg_inc_cl)`),
a `r fmtp(inc_gini_pct_red)` reduction, while the wealth Gini declines by
`r fmt2(w_gini_diff)` points on average (from `r fmt2(avg_w_hh)` to `r fmt2(avg_w_cl)`), or about
`r fmt1(w_gini_pct_red)`%. Inequality between households is not only higher than
inequality between clans on average but consistently higher across all survey years
for both income and wealth. Over time, trends in household and clan inequality are similar
for wealth, but diverge for income: clan-level income inequality has increased more
rapidly than household-level income inequality
(`r fmtp(inc_cl_pct)` vs. `r fmtp(inc_hh_pct)` since 1969). Together, these findings
demonstrate the importance of considering extended families---in addition to
households---when exploring contemporary patterns of inequality in the U.S.

# Significance Statement
Economic inequality in the United States is typically measured across households, overlooking the role of extended families in shaping inequality. Using more than five decades of data from the Panel Study of Income Dynamics, we show that income and wealth inequality are substantially lower when measured across intergenerational extended families rather than households, a finding that holds across all survey years. However, income inequality across extended families has increased more rapidly than household income inequality over time. Together, these findings demonstrate the importance of considering extended families—in addition to households—when exploring contemporary patterns of inequality.  

# Introduction
Income and wealth inequality have increased sharply in the United States from the 1960s to the present. This trend is evident in the Gini index, a widely used measure of inequality ranging from 0 (perfect equality) to 1 (perfect inequality). For instance, the U.S. household-level income Gini rose from 0.369 in 1967 to 0.418 in 2023 [@fred_gini_2025]. Similarly, the U.S. household-level wealth Gini increased from 0.79 to 0.85 from 1989 to 2019 [@barro_decline_2020; @kuhn_income_2025].

The Gini—like most measures of inequality—is typically calculated at the household level [@auten_income_2024; @killewald_wealth_2017]. As a result, research on inequality typically focuses on differences between households [@saez_wealth_2016]. This emphasis aligns with broader social-science practice, where households serve as the main unit of analysis. In doing so, scholars are largely centering nuclear families, reflecting longstanding assumptions about the primacy of the nuclear family and the diminished role of extended kin in the U.S. and other high-income industrialized nations [@goode_world_1963; @parsons_essays_1954].^[The exception is multi-generational co-residing households, which comprise about 20% of U.S. households [@pilkauskas_beyond_2018].  In this case, a focus on households elevates the importance of live-in extended family members and downplays the importance of extended family members who do not co-reside.] Data availability reinforces this orientation: most large surveys, including the Survey of Income and Program Participation and the Survey of Consumer Finances, are designed around households, limiting researchers’ ability to study inequality beyond the nuclear family.

Yet this predominant focus on households overlooks the fact that households are embedded in larger kin networks that structure obligations, cooperation, and access to resources. Such ties advantage and constrain households, shaping resource accumulation beyond parent-child transmission. Indeed, a growing body of research suggests that social scientists have underestimated the importance of extended family in studies of inequality [@killewald_wealth_2017; @mare_multigenerational_2011; @obrien_family_2024; @shiffer-sebba_keeping_2025].  Mare [@mare_multigenerational_2011], for example, argues that intergenerational mobility has likely been overstated because analyses typically ignore the influence of ancestors and extended kin. Research on elites similarly highlights the importance of extended family networks in maintaining wealth across generations. For example, O’Brien’s [@obrien_family_2024] study of elite families in Dallas shows that focusing solely on parent–child dyads obscures how these dyads are embedded within interconnected kin networks, where intermarriage among elite families is common and wealth circulates within a small number of extended lineages. Narrow attention to households therefore makes society appear more mobile than it truly is. Such insights have prompted renewed attention to the role extended family members play in the transmission of (dis)advantage. Scholars show that extended family members such as grandparents, aunts, uncles, and cousins transfer resources to individuals that impact their socioeconomic status [@adermon_dynastic_2021; @hallsten_shadow_2023; @song_shared_2019], and that these resource flows often continue even when family members live geographically far apart [@song_shared_2019].  Yet despite these insights, most macro-level studies of inequality—such as those relying on the Gini coefficient—continue to treat households, or occasionally individuals, as the primary unit of analysis. Consequently, scholars have documented patterns of inequality across households, but know little about inequality across extended families.

Drawing on scholarship showing the importance of extended families, we ask: what do income and wealth inequality in the U.S. look like when examined not only at the household level but also at the level of intergenerational extended families—what we refer to as "clans" [@shiffer-sebba_concentration_2022].  In our empirical analyses, we use data from the Panel Study of Income Dynamics (PSID), a nationally representative longitudinal survey that has followed American households and their descendants since 1968. In the PSID, households include economically interdependent individuals living in the same dwelling, while clans include all descendants linked to an original household through a family identifier. In practice, this means that clans are comprised of households linked by birth, marriage, and adoption spanning multiple generations and extended kin ties (grandparents, aunts, uncles, and cousins and in some cases great-grandparents, great-uncles, second cousins, and so on).

We calculate Gini coefficients for both income and wealth at the household and clan levels from 1969 to 2023. Conceptually, the Gini can be understood as the average difference between any two units in a population. Arithmetically, it is the average absolute difference between all pairs of observations, normalized by the mean. The Gini is often illustrated with the Lorenz curve, which plots the cumulative share of income or wealth against the cumulative share of the population—whether defined as households or clans. Perfect equality appears as a 45° line where income or wealth is distributed evenly; greater inequality pulls the curve farther below this line. The Gini summarizes this deviation from perfect equality. Comparing household- and clan-level Ginis therefore shows whether the average difference between households is larger or smaller than the average difference between clans.

Using clans as the unit of analysis requires aggregating households. One may think that aggregating units smooths variation and will inherently lead to a Gini coefficient representing less inequality. However, this is not the case. Using simulated data, we show that it is possible to obtain clan-level Gini coefficients equal to or higher than a household-level Gini coefficient (see Appendix Table A). Whether the Gini coefficient increases or decreases from the household to the clan level depends on the relationship between clan size, the overall distribution of resources, and the selection of households into clans.  

# Results
**Comparing Gini coefficients at the household and clan levels.** Figure 1 depicts income and wealth inequality for all households and clans over time. Reflecting available data, Panel A plots income Gini coefficients from 1969 to 2023 and Panel B plots wealth Gini coefficients from 1984 to 2023. Inequality is lower when using clans as the unit of analysis rather than households. The Gini coefficient for income falls by `r fmt2(inc_gini_diff)` points on average (from `r fmt2(avg_inc_hh)` to `r fmt2(avg_inc_cl)`), a `r fmt2(inc_gini_pct_red)`% decline, while the wealth Gini falls by `r fmt2(w_gini_diff)` points on average (from `r fmt2(avg_w_hh)` to `r fmt2(avg_w_cl)`), about a `r fmt1(w_gini_pct_red)`% decrease. Taken together, these results indicate that both wealth and income inequality are higher when measured using households versus extended family units.


```{r figure1, fig.width=14, fig.height=11}
grid.newpage()
grid.draw(fig1)
```

```{r figure1-note, results="asis"}
print_note(fig1_note)
```

Figure 1 shows that inequality between households is consistently higher than inequality between clans across all survey years for both income and wealth. This means that if two clans were randomly selected and compared by income or wealth in any available year, they would, on average, differ less than two randomly selected households.

We find inequality is higher for wealth than for income. This finding is consistent with prior research showing how wealth accumulates over time and generations, retaining the imprint of earlier opportunities and constraints in a way that income does not [@killewald_wealth_2017; @hallsten_grand_2017]. As such, wealth reflects long-run inequality and intergenerationally accumulated advantages and disadvantages. 

Changes over time in household and clan inequality are similar for wealth, but not for income. Income inequality measured at the household level rose by `r fmt1(inc_hh_pct)`% from `r fmt2(inc_hh_1969)` in 1969 to `r fmt2(inc_hh_2023)` in 2023. Income inequality measured at the clan level, meanwhile, rose `r fmt1(inc_cl_pct)`% from `r fmt2(inc_cl_1969)` to `r fmt2(inc_cl_2023)` over the same period. As both measurements are trending upwards, this means that clan income inequality is increasing at a faster pace than household income inequality. If we were to project these trends linearly into the future, by `r inc_convergence_yr` household and clan income inequality would appear identical. Put another way, these trends reflect a narrowing of income inequality between households and clans over time: in 1969 the gap between the two was `r fmt2(inc_gap_1969)`. By 2023 it had decreased to `r fmt2 (inc_gap_2023)`. But the gap is still far from closing.

Unlike income inequality, the gap between household and clan wealth inequality has  remained remarkably consistent. Wealth inequality typically exhibits greater temporal stability, whereas income inequality is more dynamic, because income is a flow that responds more quickly to labor-market conditions and policy shifts. Household wealth Gini has increased only slightly from `r fmt2(w_hh_1984)` in 1984 to `r fmt2(w_hh_2023)` in 2023, a `r fmt1(w_hh_pct)`% increase. At the clan level, it rose from `r fmt2(w_cl_1984)` to `r fmt2(w_cl_2023)` over the same period, which also constitutes a `r fmt1(w_cl_pct)`% increase. In other words, the clan-household gap in wealth inequality as measured by the Gini coefficient has remained stable, moving from `r fmt2(w_gap_1984)` in 1984 to `r fmt2(w_gap_2023)` in 2023. These trends should be interpreted as inequality that excludes the wealthy. The PSID does not adequately sample the top of the wealth distribution [@insolera_overview_2021], which drives much of the inequality in wealth [@insolera_overview_2021; @pfeffer_measuring_2016]. Therefore, these patterns indicate stable wealth inequality, for both households and clans, among the bottom 95% [change number depending on cites], not for society as a whole.

**Comparing household and clan distributions.** Figure 2 illustrates differences in income and wealth inequality estimates using Lorenz curves at two points in time: 1984 (blue) and 2023 (orange). These curves teach us two things about the differences between household and clan inequality: first, the differences between inequality measured through the two units are driven by different parts of the distribution for income and wealth. In both 1984 and 2023, the clan lines (dotted) separate from the household lines (solid) across the entire distribution for income (Panel A), but mostly in the middle of the distribution for wealth (Panel B). These differences, which are partially but not entirely generated by the greater prevalence of zero wealth than zero income, mean that the gap between household and clan wealth inequality are driven by those at the middle of the wealth distribution, whereas the gap between household and clan income inequality capture differences between units across the entire distribution—from those with the lowest incomes to those at the 95th percentile of the income distribution.


```{r figure2, fig.width=14, fig.height=11}
grid.newpage()
grid.draw(fig2)
```

```{r figure2-note, results="asis"}
print_note(fig2_note)
```

Second, the Lorenz curves bring into focus the different scales of the differences between household inequality and clan inequality for income and wealth. For income inequality (Panel A), the difference between household and clan inequality appears smaller compared with changes in inequality between 1984 and 2023. That both 1984 curves are closer to the line of equality than both 2023 curves means that the increase of income inequality over time is starker than differences in unit of analysis. But for wealth inequality (Panel B) this is not the case. For wealth inequality, both clan (dotted) lines are closer to the line of equality than both household (solid) lines. This means that the difference across unit of measurement is starker than the change over time.

**How the unit of analysis affects the racial wealth gap** 
Scholars document a persistent racial wealth gap between Black and non-Black households [@killewald_wealth_2017; @oliver_black_2013], while also highlighting the unique role of kinship networks in shaping economic outcomes by race [@beckert_durable_2022; @obrien_structural_2020]. Given these findings, one might expect the choice of unit of analysis to affect inequality estimates differently across racial groups. While there is higher inequality among Black households and clans overall, the direction of the household–clan difference in Gini coefficients is consistent across racial groups (see Appendix C1).

Measuring inequality at the clan-level reveals less inequality across racial subgroups. However, using clans as the unit of analysis does not eliminate racial wealth gaps between groups. Table 4 shows racial wealth gaps at both the household and clans levels for mean and median wealth. When using the mean, Black households hold `r fmt0(w_mean_ratio_hh * 100)` cents for every dollar of wealth held by non-Black households, and Black clans hold `r fmt0(w_mean_ratio_cl * 100)` cents for every dollar held by non-Black clans. A similar pattern appears when examining median wealth among clans, where Black clans hold `r fmt0(w_med_ratio_cl * 100)` cents for every dollar held by non-Black clans. However, the median reveals an even starker disparity at the household level: the typical Black household holds only `r fmt0(w_med_ratio_hh * 100)` cents for every dollar of wealth held by the typical non-Black household. These patterns suggest that while Black households hold substantially less wealth on average, wealth among them is also distributed with greater inequality. At both levels, inequality is about the same or lower between clans than between households, though Black clans hold considerably less wealth than non-Black ones.

Examining the share of households and clans with zero wealth further illustrates the racial wealth gap at the household and clan levels and how this inequality appears lesser at the clan-level. Following [@oliver_black_2013], `r fmtp(zero_w_hh_black)` of Black households report no wealth, compared to just `r fmtp(zero_w_hh_nonblack)` of non-Black households. At the clan level, however, the share with zero wealth drops considerably. Non-Black clans exhibit extremely low levels of zero wealth, with only `r fmtp(zero_w_cl_nonblack)` reporting none, suggesting that most non-Black individuals have some wealth within their extended family networks. Black clans also experience a decline in zero-wealth, falling to `r fmtp(zero_w_cl_black)`. In percentage terms, the transition from the household to clan level represents a `r fmt1(zero_w_pct_red_black)`% reduction in zero-wealth for Black households and a `r fmt1(zero_w_pct_red_nonblack)`% reduction for non-Black households. As a result, the extent of inequality appears less stark when wealth is considered at the clan level rather than at the household level, though the decrease in inequality is higher for non-Black clans. 

```{r race-table-wealth}
race_wealth_ft
```

```{r race-table-note, results="asis"}
print_note(fig4_table_note)
```



*We are really talking about the racial wealth, not income, gap. It would probably make sense to exclude the table below but it is here in case you want to see it. The race ratio for clans is **worse** than it is for households for income.*

```{r race-table-income}
race_income_ft
```

*Below are two figures that show the information in the tables. These (or a combination of these) could be used instead of tables. I (AEB) found the tables easier to interpret, so that's what I took a stab at first.*

```{r figure3, out.width="\\linewidth"}
knitr::include_graphics(here::here("9_figures","output","figure3.pdf"))
```

```{r figure3-note, results="asis"}
print_note(fig3_note)
```

```{r figure4, out.width="\\linewidth"}
knitr::include_graphics(here::here("9_figures","output","figure4.pdf"))
```

```{r figure4-note, results="asis"}
print_note(fig4_note)
```


**Sensitivity of difference between household- and clan-level Gini coefficients.** Across all specifications, estimated inequality is lower when using clans rather than households as the unit of analysis. This finding is robust across race, different inclusion criteria for income and wealth measures, different samples, alternative measures of inequality, and when adjusting for household and clan size. Results from sensitivity analyses are available in Appendix C and D. *This mini-section should change or be deleted.*


# Discussion
Our analyses explored what income and wealth inequality in the U.S. look like when examined not only at the household level but also at the level of intergenerational extended families.  We found that both income and wealth inequality were more extreme, as per Gini coefficients, when examining society as a collection of households (the standard in the literature) rather than a collection of clans.  Income inequality was about 6 points higher for households than for clans, and wealth inequality was about 12 points higher, indicating fewer disparities on average between extended family networks than between households. In other words, an individual’s extended family was more economically similar on average to other extended families than their household was to other households. Thus, claims that current estimates, which focus on households, underestimate inequality [@mare_multigenerational_2011] may be exaggerated. Inequality between larger family units is lower than inequality between households. 

Our analyses provided an important complement to existing scholarship on income and wealth inequality that typically measures inequality at the household level (or in some cases the individual level). Our aim is not to downplay the importance of households as units of analysis, but rather to show that considering multiple units of analysis—including households and clans—is helpful for fully understanding inequality as a complex societal phenomenon. The choice of analytical unit matters in part because different historical periods generate changes in inequality at some levels more than others. For example, our analyses showed that clan-level income inequality has increased more rapidly than household-level income inequality over the study period. By contrast, trends in household and clan inequality were similar for wealth over time. This may reflect the cumulative nature of wealth, which is more durably concentrated within kinship networks [@obrien_family_2024; @adermon_dynastic_2021; @pfeffer_determinants_2017], or limitations of the PSID’s coverage of the top of the wealth distribution, where much wealth inequality is generated. Together, these patterns reinforce the importance of examining inequality at multiple levels of social organization.

Differences in historical patterns between households and clans may reflect changes in mobility over time. Mechanically, lower mobility produces greater homogeneity within clans over time, whereas higher mobility has the opposite effect. Conversely, inequality across extended families may build atop inequality across households in reducing mobility. If resources and constraints are transmitted within clans then rising inequality may further reduce mobility for those in disadvantaged households. In other words, the prevalent concern that high inequality undermines intergenerational mobility [@corak_income_2013; @ermisch_parents_2012; @jerrim_income_2015] may operate on both the household and clan levels, fueling future differences in inequality between households and clans. Although our analyses focused on the macro level, they highlighted a clear need for additional research on how extended families structure social life at the micro level. One key question raised by our findings concerns the prevalence and scale of resource transfers within extended families, an issue beyond the scope of this study. Methodologically, our project suggests the need for data and research designs that capture resource flows across kinship networks rather than limiting analysis to households. Taken together, our findings demonstrate that understanding economic inequality requires moving beyond a sole focus on households to also account for the extended family structures through which resources, advantage, and disadvantage are transmitted across generations.

# Materials and Methods
**Data and sample.** We use the genealogical design of the Panel Study of Income Dynamics (PSID) to compare estimates of inequality when using households and clans as the unit of analysis. The PSID began in 1968, surveying families annually until 1997 and biennially thereafter. It follows descendants of the original sample across generations and maintains national representativeness. Data from the PSID produce estimates that mostly align with other national surveys [@insolera_overview_2021]. 

The PSID facilitates two linked units of analysis: households and clans. Households consist of economically interdependent individuals living in the same dwelling, with one member reporting data on income and wealth for all members. Household composition changes over time with transitions such as divorce or childbirth; individuals who leave receive a new household identifier. As a result, unique households cannot be identified over time in the PSID. In contrast, clans include all descendants of households sampled in 1968, linked through a PSID-generated identifier, which we refer to as a clan identifier. Households added after 1968 receive their own clan identifier upon entry. Unlike household identifiers, clan membership remains fixed for individuals and can be compared over time. A small number of households (`r fmt0(n_mismatched)` household-years) contain couples belonging to different clans; these cases were assigned to both clans. 

Additionally, although all households belong to a clan, not all extended family members are observed in the PSID in every year. In some cases (especially in earlier years)^[The prevalence of single-household clans dropped by 75% between 1969 and 1980. However, overall trends remain similar regardless of whether this earlier period is excluded.], PSID clans consist of only a single observed household, rendering household and clan-level measures identical. These observations are excluded from the main analyses because they do not capture extended family dynamics and are a result of survey limitations, rather than social reality. Including these single household clans reduces the household–clan inequality gap for income and wealth but does not eliminate it (see Appendix C2).

The PSID has collected income data since its inception but began collecting wealth data only in 1984. As a result, analytic samples differ: income analyses include `r fmt0(inc_hh_years)` household-years and `r fmt0(inc_cl_years)` clan-years, representing `r fmt0 (inc_uniq_clans)` unique clans; wealth analyses include `r fmt0(w_hh_years)` household-years and `r fmt0(w_cl_years)` clan-years, representing `r fmt0(w_uniq_clans)` unique clans.

**Measures of income and wealth.** We measure both income and wealth to capture distinct dimensions of economic inequality. This choice follows from research identifying divergent patterns depending on choice of financial indicator because mechanisms driving wealth inequality are distinct from those driving income inequality [@killewald_wealth_2017; @pfeffer_determinants_2017; @hallsten_grand_2017]. Aggregate measures for income and wealth are provided by the PSID. Income includes earnings, asset income, transfers, and income from farming and gardening. Wealth includes non-primary real estate, business equity, stocks, bonds, vehicles, retirement accounts, and bank accounts, and includes home equity. As a sensitivity analysis, we compare Gini coefficients for wealth measures that exclude home equity (see Appendix C3). When home equity  is excluded, estimated inequality is lower at both the household and clan level, but the difference between household and clan-level Gini coefficients remains similar.

Missing data for components of income and wealth are imputed by the PSID. Concerns about measurement error, especially for wealth, are minimal, as imputation methods utilized by the PSID explain little of the variation in wealth over time [@pfeffer_determinants_2017; @si_multiple_2021]. All income and wealth values are adjusted for inflation. Additionally, analyses set negative values of income and wealth to zero so that Gini coefficients are interpretable by remaining in the 0 to 1 range. When negative values are included, Gini coefficents are about the same (see Appendix C4). 

To construct clan-level measures, we aggregate household income and wealth for all households in a clan within each clan-year. Measures are divided by the number of people in a household or number of households within a clan to standardize by household and clan size, respectively. Adjusting for size descreased inequality and magnifies the difference between household and clan Gini coefficients. Nonetheless, less inequality is estimated when clans are the unit of analysis whether income and wealth are adjusted for size or not (see Appendix C5).

All results presented are weighted. Using weights does not meaningfully change results (see Appendix C6). Household-level weights are provided by the PSID. The PSID constructs family weights from individual weights using the fair shares method, which assumes that the probability of observing any individual within a family equals the probability of observing the family itself. We extend this approach to construct clan-level weights using the same fair shares principle. Clan weights assume that the probability of observing a family within a clan equals the probability of observing the clan itself. As a result, each clan’s influence reflects the representation of the average household within that clan, rather than scaling mechanically with the clan’s total number of households. 

**Measures of race.** The PSID collects race data from household heads and, in some years, from spouses. If a respondent ever identifies as Black, then that value is imputed for all years that respondent appears in the data [@sayer_women_2004]. Black households are all households with a Black head of household. Black clans are clans in which more than half of the households within it identify as Black. Analyses by race and ethnicity are limited to Black and non-Black categories because there are relatively few clans in the PSID data (`r fmt0 (inc_uniq_clans)` for income and `r fmt0(w_uniq_clans)` for wealth). Additionally, the genealogical design of the PSID reflects the demographics of the United States at the time the survey began in 1968, which constrains analyses of clans for demographic groups that were not prominent at that time. Although the PSID incorporated additional samples over time to reflect the changing demographics of the United States in the PSID sample, ethnicity analyses remain limited because Hispanic/Latino households do not form consistent multigenerational lineages over the full panel and therefore have limited clans.

**Indexing inequality with the Gini coefficient.** Inequality is measured using the standard Gini coefficient, which ranges from 0 (perfect equality) to 1 (perfect inequality). While the Gini index is a widely-used measure of inequality, it has well-known limitations. First, the Gini places greater weight on differences around the middle of the income or wealth distribution, making it well suited to the PSID, which underrepresents wealthy households compared with surveys such as the Survey of Consumer Finance [@atkinson_measurement_1970; @de_maio_income_2007; @kawachi_relationship_1997]. To assess whether our results depend on where inequality is concentrated along the distribution, we also examine two related measures, referred to as the Gini’s “nuclear family” [@aaberge_ginis_2007]. Across all measures, we find that estimated inequality is consistently lower when measured at the clan level than at the household level for both income and wealth (see Appendix G). 

Gini coefficients are calculated separately for income and wealth to reflect their distinct dimensions of inequality. Estimates of Gini coefficients in our data are comparable to estimates that other studies have found, though there is more variation in wealth coefficients, likely because the PSID excludes wealthy households (see Appendix B).


# Appendix

```{r appendix-load}
source(here::here("functions", "all_functions.R"))

# Source main figures and appendix figures 
SAVE_FILES <- FALSE
source(here("9_figures", "src", "figure_main.R"))
source(here("9_figures", "src", "figure_appendix.R"))

income          <- read_gini(here("6_calculate_gini", "output", "income.csv"))
wealth_nohouse  <- read_gini(here("6_calculate_gini", "output", "wealth_nohouse.csv"))
wealth_withhome <- read_gini(here("6_calculate_gini", "output", "wealth_withhome.csv"))

income_all <- read_csv(here("6_calculate_gini", "output", "income.csv"), show_col_types = FALSE) %>%
  mutate(year = suppressWarnings(as.integer(as.character(year))))
wealth_all <- read_csv(here("6_calculate_gini", "output", "wealth_withhome.csv"), show_col_types = FALSE) %>%
  mutate(year = suppressWarnings(as.integer(as.character(year))))

inc_c123 <- read_csv(here("8_nuclear_family", "output", "income_C123.csv"),  show_col_types = FALSE)
wnh_c123 <- read_csv(here("8_nuclear_family", "output", "wealth_C123.csv"),  show_col_types = FALSE)
```

## Appendix A: Simulating Gini Coefficients

Appendix A illustrates extreme cases from simulated household and clan data. The simulation demonstrates that clan-level Gini coefficients are not necessarily lower than household-level Gini coefficients. The direction of the difference depends on the relationship between clan size, the overall distribution of resources, and the selection of households into clans. Appendix A1 shows the minimum possible difference between clan and household Gini coefficients in the simulated data, while Appendix A2 shows the maximum possible difference. Curves show that Gini coefficients at the clan level can exceed those of households. Appendix A3 illustrates that, while it is more likely for Gini coefficients at the household level to exceed Gini coefficients at the clan level, whether this effect is observed depends on the combination between clan size and income or wealth distribution.

```{r appendix-a-sim}
set.seed(123)
sims   <- run_many_sims_sums(
  n_sims = 100, n_clans = 5, hh_per_clan = 10,
  dist = "lognormal", dist_params = list(meanlog = 0, sdlog = 1),
  clan_size_scenario = "rich_big"
)
df_res <- results_from_sims_sums(sims)
ext    <- extreme_indices_sums(df_res)
```

\newpage

**Appendix A1: Case with lowest simulated clan-household Gini difference**
```{r appendix-a1, fig.width=6, fig.height=4.5, fig.align="center"}
plot_lorenz_for_sim_sums(sims[[ext$idx_min_sums]])
```

**Appendix A2: Case with highest simulated clan-household Gini difference**
```{r appendix-a2, fig.width=6, fig.height=4.5, fig.align="center"}
plot_lorenz_for_sim_sums(sims[[ext$idx_max_sums]])
```

\newpage

**Appendix A3: Summary of extreme cases**
```{r appendix-a3, fig.width=6.5, fig.height=5, fig.align="center"}
make_two_mekko_sums(sims, ext$idx_min_sums, ext$idx_max_sums)
```

\newpage

## Appendix B: Comparison with External Estimates

Appendix B compares Gini coefficients at the household level calculated for income and wealth with Gini coefficients from other researchers. Gini coefficients in Appendix B are calculated using the full sample of households in the PSID, rather than just the households who belong to a clan with at least one other household (which are the results presented in the main paper).

Gini coefficients are not available for all years in other sources, so only years available are presented. Gini coefficient benchmarks for income rely on estimates from the Federal Reserve Bank of St. Louis (FRED) and the U.S. Census [@fred_gini_2025; @noauthor_table_nodate]. FRED uses data from the World Bank on U.S. households and the Census uses data from the Current Population Survey (CPS). Gini coefficient benchmarks for wealth rely on estimates from research papers based on data from the PSID or Survey of Consumer Finances (SCF) [@wolff_size_209; @pfeffer_wealth_2013].

```{r appendix-b1}
b1_years <- c(1969L, 1979L, 1989L, 1999L, 2009L, 2019L)

b1 <- tibble(
  Year          = c("1969", "1979", "1989", "1999", "2009", "2019"),
  Year_pick     = b1_years,
  FRED          = c(0.391, 0.404, 0.431, 0.458, 0.468, 0.484),
  `U.S. Census` = c(0.349, 0.365, 0.401, 0.429, 0.443, 0.454)
) %>%
  left_join(
    income_all %>% filter(year %in% b1_years) %>%
      select(Year_pick = year, `Our Results (All HH)` = hh_w_inc),
    by = "Year_pick"
  ) %>%
  select(Year, FRED, `U.S. Census`, `Our Results (All HH)`) %>%
  mutate(across(-Year, ~ round(as.numeric(.x), 3)))

flextable(b1) %>%
  set_caption(caption = "Appendix B1. Comparison of income Gini coefficients",
              autonum = FALSE) %>%
  set_header_labels(`Our Results (All HH)` = "Our Results\n(All HH)") %>%
  theme_vanilla() %>%
  bold(part = "header") %>%
  align(align = "center", part = "all") %>%
  fontsize(size = 9, part = "all") %>%
  fontsize(size = 8, part = "body") %>%
  fit_to_width(max_width = 6.5)
```

```{r appendix-b2}
b2 <- tibble(
  Year      = c("1983", "1984", "1989", "1992", "1994", "2003", "2004", "2007", "2009", "2010", "2011"),
  Year_pick = c(NA, 1984L, 1989L, NA, 1994L, 2003L, NA, 2007L, 2009L, NA, 2011L),
  SCF       = c(0.760, NA,    0.781, 0.770, NA,    NA,    0.809, NA,    NA,    0.846, NA),
  PSID      = c(NA,    0.739, 0.742, NA,    0.732, 0.810, NA,    0.830, 0.890, NA,    0.880)
) %>%
  left_join(
    wealth_all %>%
      filter(year %in% c(1984L, 1989L, 1994L, 2003L, 2007L, 2009L, 2011L)) %>%
      select(Year_pick = year, `Our Results` = hh_w_wealth) %>%
      mutate(`Our Results` = round(as.numeric(`Our Results`), 3)),
    by = "Year_pick"
  ) %>%
  select(Year, SCF, PSID, `Our Results`)

flextable(b2) %>%
  set_caption(caption = "Appendix Table B2. Comparison of wealth Gini coefficients with external estimates",
              autonum = FALSE) %>%
  theme_vanilla() %>%
  bold(part = "header") %>%
  align(align = "center", part = "all") %>%
  fontsize(size = 9, part = "all") %>%
  fontsize(size = 8, part = "body") %>%
  fit_to_width(max_width = 6.5)
```


# Appendix C: Sensitivity Analyses — Gini Coefficients {-}

## Appendix C1: Gini Coefficients by Race Subgroup {-}
Graphs on the left side show Gini cofficients for Black households and clans. Graphs in the middle show Gini coefficients for non-Black households and clans. The graphs on the right side show the difference between household and clan (HH - Clan) Gini coefficients for each group. Both Black and non-Black clans have higher Gini coefficients than Black and non-Black households, indicating that inequaltiy is lower when estimated at the clan level. The differences between Gini coefficients at the household and clan level are not driven by racial subgroups.

```{r appendix-c1, fig.width=14, fig.height=11, out.width="100%"}
sensitivity_gini_6
```


## Appendix C2: Sensitivity — Single-Household Clans {-}
In reality, all households belong to clans that consist of multiple households. However, due to the original PSID sample lacking clans by design, as well as attrition, some clans in the PSID comprise a single household in some years. Because this limitation is imposed by the survey design, not social reality, we exclude household-years where households are alone in their clans from both household and clan measurements. The graphs on the left side show this. When including these single-household clans, or “clanless” households, the difference in Gini coefficients at the household and clans levels is minimized but not eliminated. The graphs in the middle show Gini coefficients when single-household clans are included. The graphs on the right side show the difference between household and clan (HH - Clan) Gini coefficients for each group.

```{r appendix-c2, fig.width=14, fig.height=11, out.width="100%"}
sensitivity_gini_3
```


## Appendix C3: Sensitivity — Wealth With and Without Home Equity {-}
Results presented in the main text include home equity for measures of wealth. Appendix C3 shows Gini coefficients when home equity is excluded (middle graphs), and report the difference between Gini coefficients at the household and clan levels (graphs on the right). Excluding home equity makes estimated inequality lower and minimizes the observed difference between household and clan Gini coefficients.

```{r appendix-c3, fig.width=14, fig.height=7, out.width="100%"}
sensitivity_gini_2
```


## Appendix C4: Sensitivity — Negative Values {-}
Results presented in the main text exclude negative values for income and wealth. Appendix C4 show Gini coefficients when negative values are included (middle graphs), and report the difference between Gini coefficients at the household and clan levels (graphs on the right).  Including negative values does not change estimated Gini coefficients at the household or clan levels.
```{r appendix-c4, fig.width=14, fig.height=11, out.width="100%"}
sensitivity_gini_1
```


## Appendix C5: Sensitivity — Size Standardization {-}
Results in the main text standardize by household and clan size. Estimated inequality is higher when results are not standardized by size and the gap between household and clan Gini coefficients is smaller. However, Gini coefficients are higher at the clan versus household level irrespective of whether measurements are adjusted by size. 

```{r appendix-c5, fig.width=14, fig.height=11, out.width="100%"}
sensitivity_gini_5
```


## Appendix C6: Sensitivity — Weighting {-}
All results presented in the main paper are weighted. Household-level Gini coefficients are weighted using PSID-provided family weights, while clan-level Gini coefficients are weighted using a constructed clan weight based on the PSID family weight. Weighting does not substantially change the results.

```{r appendix-c6, fig.width=14, fig.height=11, out.width="100%"}
sensitivity_gini_4
```


# Appendix D: Alternative Inequality Measures {-}
Appendix D shows three alternative measures of inequality proposed by Aaberge [@aaberge_ginis_2007]. C1 is the Bonferroni coefficient (greater weight on lower tail); C2 is the standard Gini coefficient; C3 places greater weight on the upper tail of the distribution. All estimates are weighted averages across all available years. Inequality is lower at the clan level using any measure of inequality. 

```{r appendix-d, fig.width=14, fig.height=13, out.width="100%"}
appendix_c
```


